---
title: "Lecture 1 code"
author: "Kevin Wang"
date: "March 7, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}

#Create a code chunk called setup.
#Do not include the code of this chunk or it's output in the final document.

knitr::opts_chunk$set(echo = TRUE) #by default, print the code in the document.

```


# Lecture 1

## Read in data

Let's start by reading in some data...


```{r readData}

#read in data
data <- read.csv("2016Fatalities.csv",header=T) 

#Try the following too
#data <- read.delim("2016Fatalities.csv",header=T,sep = ",") 


```


## Exploration

Now lets get a feel for what our data looks like... 

```{r explore}

#get dimension of data
dim(data)

names(data) #Lists all the variables
colnames(data)  #Lists all the variables

data[1,]  #Extracts the 1st row
head(data)  #List the 1st 6 rows of data, tail() the last 6

class(data)  #Shows the way R has stored the data

str(data)

CLASS = NULL
for(i in colnames(data)) CLASS[i] = class(data[,i]) 
CLASS  #shows how the data are stored in each column of the dataframe
```


## Plots


Have a look at Days of the week variable using a frequency table and ordinal plot.
```{r dayofweek}
DayWeek  =  data$Dayweek  #Extracts day of the week
table(DayWeek)
plot(table(DayWeek),las=2)
```


While we're at it, check out speed limit.
```{r speedLimit}
Speed <- data$SpeedLimit  #Extracts SpeedLimit
table(Speed)
plot(table(Speed))
```



We can generate a histogram of Age
```{r Histograms}
Age = data$Age
min(Age)
max(Age)

hist(Age,xlab="Age",
     main="Regular Histogram for Age of Fatality")

breaks = c(-10,18,25,70,100) # This creates a vector that we will use to define the breaks
table(cut(Age,breaks,right=F))   
hist(Age,br=breaks,freq=F,right=F,
     xlab="Age",
   main="Probability Histogram for Age of Fatality")  


hist(Age,br=breaks,freq=T, right=F, main ="Misleading Regular Histogram")
```



## Summation notation

The answers to the 'try it yourself' page on summation notation.
```{r summationNotation}
#Check your answers
x=c(1,4,6,2,3,7)
y=c(sum(x), sum(x[2:5]^2), sum(c(1:6)*x), sum(sort(x)-1))
y

```


## Mean and median

We can find the mean and median using the following...

```{r meanMedian}
#Check your answers
x=c(1,4,6,2,3,7)
mean(x)
median(x)
```


## Skewness

```{r skew}

# Set a seed to reproduce "random sampling" in the future
set.seed(123)

# Generate some data... We will learn about how to do this in two weeks
x1=rnorm(1000)
x2=-rgamma(1000,1,2)
x3=rgamma(1000,1,2)


#Make three histograms

par(mfrow = c(3, 1)) #Creates a plotting window with 3 rows and 1 columns
hist(x1,main="Symmetric",breaks = 20)
abline(v = c(mean(x1),median(x1)),col = c(2,4)) # Draw vertical lines where the mean and median are that are red(2) and blue(4)... could use col=c('red','blue') instead.
hist(x2,main="Left skewed",breaks = 20)
abline(v = c(mean(x2),median(x2)),col = c(2,4))
hist(x3,main="Right skewed",breaks = 20)
abline(v = c(mean(x3),median(x3)),col = c(2,4))
par(mfrow = c(1, 1))
```

We'll look at speed limit again...
```{r Speed2}
Speed <- data$SpeedLimit
mean(Speed)
median(Speed)
hist(Speed,breaks = 20)
abline(v = c(mean(Speed),median(Speed)),col = c(2,4))

```

## Multimodal data

Data can have multiple modes...
```{r Modality}
x1=rnorm(1000)
x2=c(rnorm(1000,25,6),rnorm(1000,100,4))
x3=c(rnorm(1000,25,6),rnorm(1000,100,4), rnorm(1000,200,16))
par(mfrow = c(1, 3))
hist(x1,main="Unimodal")
hist(x2,main="Bimodal")
hist(x3,main="Trimodal")
```


## Spread

Data can be centered in the same spot but have different spreads. 

```{r spread}
x1=rnorm(1000,0,1)
x2=rnorm(1000,0,2)
par(mfrow = c(1, 2)) 
hist(x1,xlim = c(-6.5,6.5))
hist(x2,xlim = c(-6.5,6.5))
```


We can calculate a few measures of spread.
```{r robustSpread}

x =  c(1,4,6,2,3,7)
sd(x) #standard deviation of x

fivenum(x) # The quantiles and min,max of x

x2=c(1,4,6,2,3,7,8)
fivenum(x2)


fivenum(x)[4]-fivenum(x)[2] #IQR of x

# IQR is more robust than sd
x1=c(1,4,6,2,3,7,100)
sd(x1)
fivenum(x1)[4] - fivenum(x1)[2]
```


## Boxplots

We can use boxplots to visualise the 5 number summary.
```{r boxplot}

fivenum(x)
boxplot(x, horizontal=T, col="purple")

#Boxplot of Age
boxplot(Age,horizontal=T)


#Split by men and women
AgeM <- data$Age[ data$Gender == "Male"]
AgeF <- data$Age[ data$Gender == "Female"]
par(mfrow = c(1, 2))  #Puts 2 boxplots in a row
boxplot(AgeM)
boxplot(AgeF)

par(mfrow = c(1, 1))
#Alternatvely
boxplot(Age~data$Gender, ylab="Age")

#Or...
boxplot(AgeM,AgeF, ylab="Age")


```

Boxplots can show us weird things about data
```{r}
boxplot(data$SpeedLimit, horizontal = T)
```


## Outliers

We can identify outliers in the data by using arbitrary definitions
```{r}

Usyd <- read.csv("USyd.csv")
heights1=c(Usyd$Heights[1:27],178) #add in an outlier
iqr=fivenum(heights1)[4]-fivenum(heights1)[2]
lt=fivenum(heights1)[2]-1.5*iqr
ut=fivenum(heights1)[4]+1.5*iqr

# 1.5 * IQR rule
heights1[(heights1<lt) | (heights1 > ut)]   # | = 'or'

# 3 * SD rule
3*sd(heights1)  
heights1[abs(heights1-mean(heights1))>3*sd(heights1)]
```


```{r transformation}
w=c(1,2,3,4,10,30,60,120,180,300)
w1=log(w,10)
par(mfrow = c(1, 2))
boxplot(w, main ="Data")
boxplot(w1, main="Log of Data")
```

